---
title: "Leydet_Lab03"
author: "David Leydet"
date: "2022-09-07"
output: html_document
---

GEOG6000 Methods of Data Analysis Lab 03

Experiment using R Markup in HTML

Plotting Types
```{r}
setwd("~/Desktop/University of Utah PhD /Course Work/Fall 2022 Semester/GEOG6000_Data Analysis/lab03") ##Set working directory

##Bar Plots

VAdeaths = read.csv("../datafiles/VAdeaths.csv", row.names = 1)
mycol = heat.colors(5)
barplot(as.matrix(VAdeaths), beside=T, legend = rownames(VAdeaths),
        col = mycol,
        main = "VA Deaths by Age",
        ylab = "Number of Deaths",
        xlab = "Location and Sex")
```

Dot Charts!

```{r}

VAdeaths.m = as.matrix(VAdeaths)
dotchart(VAdeaths.m)

```
```{r}
##Transpose the matrix --> Ask about this in lab...I am still having difficulty conceptualizing the data organizing coding we are using***

##Data Structures
#Data can be numeric, factor, character, or boolean
#Vector is a set a values in one dimension
#Matrix is a 2 dimensional type rows/columns
#Array is like a matrix with more then 2 rows/columns
#Data Frame like csv file
##'as' functions convert data types to use for the function, i.e. as.matrix

dotchart(t(VAdeaths.m)) 

##Transpose for rural/urban category to the age bracket##

```

Plotting Multiple Series with Points

```{r}

cmp = c(52, 57, 62, 67, 72)
plot (cmp, VAdeaths$Rural.Male, pch = 1, col= 1, ylim = c(0,70),
      xlab = "Age Class",
      ylab = "Mortalitiy",
      main = " Virginia Death Rates")
points(cmp, VAdeaths$Rural.Female, pch = 2, col = 2)
points(cmp, VAdeaths$Urban.Male, pch = 3, col = 3)
points(cmp, VAdeaths$Urban.Female, pch = 4, col = 4)
legend("topleft", legend = c("Rural Male", "Rural Female", "Urban Male", "Urban Female"),
       col = c(1, 2, 3, 4), pch = c(1,2,3,4))
##This code concatenated the each group to its own symbol and color on the plot and legend.
##1, 2, 3, 4 was in order of rural male, rural female, urban male, and urban female
```

Plotting Multiple Series with Lines

```{r}

ipcc = read.csv("../datafiles/ipccScenario_1900_2100.csv")
plot(ipcc$yrs, ipcc$commitMed, 
     type = 'l',
     lwd = 1,
     col = 'orange',
     ylim = c(-1.0,3.5), 
     main = 'IPCC Scenarios', 
     xlab = 'Years',
     ylab = 'Global Temp.')
lines(ipcc$yrs, ipcc$b1Med, lwd = 1, col = "blue")
lines(ipcc$yrs, ipcc$a1bMed, lwd = 1, col = 'green')
lines(ipcc$yrs, ipcc$a2Med, lwd = 1, col = 'red')
###This plot takes multiple data and plots them against a common x axis which is in the first part of the plot and line argument

abline(h = 0, lty = 2)
legend("topleft", legend = c("Commit", "B1", "A1B", "A2"), 
       lty = 1,
       lwd = 1,
       col = c('orange', 'blue', 'green', 'red'))

##Experiment with elements of the chart to make it more visually appealing
```

Plotting Polygons

```{r}

plot(ipcc$yrs, ipcc$commitMed, 
     type = 'n', 
     ylim = c(-1.5,1.0), 
     main = 'Commit Scenario',
     xlab = 'Years',
     ylab = 'Global Temp.')
##This produces a "blank plot with an x and y axis along with labels

polygon(c(ipcc$yrs,rev(ipcc$yrs)),c(ipcc$commitLo,rev(ipcc$commitHi)), 
        col = 'orange')
##How does the "rev" function work??****
##Need to reverse it in order to give it its total set of vertices instead of going back to the beginning (think drawing as opposed to typewriter)

lines(ipcc$yrs, ipcc$commitMed, 
      lwd = 2, 
      col = 'black')

```

Polygon Fill with Shading Lines 

```{r}

plot(ipcc$yrs, ipcc$commitMed, 
     type = 'n', 
     ylim = c(-1.5,1.0))

polygon(c(ipcc$yrs,rev(ipcc$yrs)),c(ipcc$commitLo,rev(ipcc$commitHi)), 
    col = 'orange', 
    density = 20, 
    angle = 145)
lines(ipcc$yrs, ipcc$commitMed, 
      lwd = 2, 
      col = 'black')

```

Plotting Images 

```{r}

volcano = read.table("../datafiles/volcanodem.txt")
volcano = as.matrix(volcano)
##Where do the the V"n" columns come from? Matrix transition?

z = 2 * volcano ##Exaggerate the relief
x = 10 * (1:nrow(z))
y = 10 * (1:ncol(z))

##Not sure how we are organizing the data here.

image(x, y, z)

```

Image with better color scheme:
```{r}

image(x, y, z, col = terrain.colors(100))

```

Contour Plot:

```{r}
contour(x, y, z, nlevel = 20)

##Can this be used for isopleths???

```

Perspective Plot:
```{r}

persp(x, y, z,theta = 210, phi = 15, scale = FALSE)

##Neat!

```

Shaded Perspective Plot

```{r}

persp(x, y, z, 
      theta = 130, 
      phi = 30, 
      scale = FALSE, 
      col = 'green3', 
      ltheta = -120, 
      shade = 0.75, 
      border = NA, 
      box = FALSE)

##ltheta controls the light angle
##shade controls the diffusion of the lighting 
#phi is the vertical angle
#theta is the horizontal view angle

```


```{r}
pdf('volcano.pdf')
image(x, y, z, 
      col = terrain.colors(100), 
      main = "Maunga Whau DEM")
dev.off()

##Export to .pdf
```


Simple Linear Regression

```{r}

regrex = read.csv("../datafiles/regrex.csv")
summary(regrex)


```


```{r}
plot( y ~ x, data=regrex)

```

```{r}
cor.test(regrex$x, regrex$y)
```

```{r}

ex1.lm = lm( y ~ x, data = regrex)
ex1.lm

```
```{r}

anova(ex1.lm)
##Is this appropriate given only two variables?

```

```{r}

summary(ex1.lm)

```
This model seems to be a good fit based on the residuals distribution, p-value, and adjusted r-squared

Plotting the regression with a line
```{r}

plot( y ~ x, data = regrex,
      main = "Linear Regression Example",
      xlab = "X",
      ylab = "Y",
      col = "blue",
      pch = 2)
abline(ex1.lm, col = "red")

```

Model Standard Errors
```{r}
predict(ex1.lm, level = 0.95, interval = "conf")

##What am I looking at?
##Fit = predicted value
##lwr = lower interval
##upr = upper interval -> In this case 95%?
```

```{r}
newx = data.frame(x = -1:30) ##New data frame with values x from -1 to 30

newy = predict(ex1.lm, newdata = newx,
               level = 0.95,
               interval = "conf") ##prediction of our new y values with our new data newx

str(newy)

plot( y ~ x, data=regrex, 
    pch = 16,
    main = 'Model Confidence Intervals')

lines(newx$x, newy[,"fit"], col = 2) ##predicted values of x...or fit?
lines(newx$x, newy[,"lwr"], col = 3, lty = 2) ##lower confidence interval?
lines(newx$x, newy[,"upr"], col = 3, lty = 2) ##upper confidence interval?

```

Residual Plots

```{r}

ex1.res = residuals (ex1.lm) ##creating a variable for the residuals
hist(ex1.res)

```

```{r}

plot(ex1.lm, which = 1)

##Question on this

```

```{r}

plot(ex1.lm, which = 2)

```

```{r}

plot(ex1.lm, which = 4)

```

```{r}

shapiro.test(ex1.res)

#The null hypothesis is that the data are normally distributed. Here we get a high p-value, and no evidence to reject the null.

```

GG PLot Example During Lab (September 8, 2022)

```{r}
library(ggplot2)
library(ggthemes)

head(ipcc)

##need to put it into long format data

plot_df = data.frame(
  yrs = rep(ipcc$yrs, 4), #replicate 4 columns by year
  cilo = c(ipcc$commitLo, ipcc$b1Lo, ipcc$a1bLo, ipcc$a2Lo),
  cihi = c(ipcc$commitHi, ipcc$b1Hi, ipcc$a1bHi, ipcc$a2Hi),
  med = c(ipcc$commitMed, ipcc$b1Med, ipcc$a1bMed, ipcc$a2Med),
  scenario = rep(c('Commit', 'b1', 'a1b', 'a2'), each = nrow(ipcc))
)

plot_df

ggplot(plot_df, aes(x = yrs, y=med, col = scenario)) +
  geom_line(size = 1.25) +
  theme_fivethirtyeight()

```

Simon's Code

```{r}

a2_ipcc <- subset(plot_df, scenario == "a2")
ggplot(a2_ipcc, aes(x = yrs)) +
  geom_ribbon(aes(ymin = cilo, ymax = cihi), alpha = 0.5, 
              col = NA, fill = 'skyblue') +
  geom_line(aes(y = med), size = 1.25) +
  theme_bw()

ggplot(plot_df, aes(x = yrs, col = scenario, fill = scenario)) +
  geom_ribbon(aes(ymin = cilo, ymax = cihi), alpha = 0.25) +
  geom_line(aes(y = med), size = 1.25) +
  theme_bw()

ggplot(plot_df, aes(x = yrs, col = scenario, fill = scenario)) +
  geom_ribbon(aes(ymin = cilo, ymax = cihi), alpha = 0.25) +
  geom_line(aes(y = med), size = 1.25) +
  theme_bw() +
  facet_wrap(~ scenario)



```

```{r}
##Model Confidence Interval versus Prediction Model Confidence Interval

##We are 95% sure that the model line will fall between the confidence interval lines

##Prediction interval 




```









